Predicting Remaining Useful Life of NASA Aircraft Engines

### 2025-03-09 index.qmd
3/8 updates

To determine whether to use CPH for modeling, the data must contain a
time to event value and meet various assumptions, including proportional
hazards. Data that is censored will require a survival analysis that
type of analysis is deliberately designed for censored data. The data
must also have a binary indicator variable for the event occurrence and
no missing values [@nahhas2025].

### Equations

**General Equation**

CPH can be estimated as

$$ h(t|\mathbf{Z}) = h_0(t)\text{exp}(\sum_{k=1}^{p} \beta_kZ_k) $$

-   $h(t|\mathbf{Z})$ is the hazard rate at time $t$ for an individual
    with risk vector $\mathbf{Z}$
-   $h_0(t)$ is an arbitrary baseline hazard rate
-   $\mathbf{Z}$ is the set of $p$ covariates
-   $\beta$ is the set of $p$ coefficients that measure the impact of
    the covariates [@klein2005]

This equation can then be adapted for the specific dataset being
studied.

**Hazards Ratio**

CPH is a proportional hazards model because the ratio of hazard rates
between two individuals is

$$ \frac{h(t|\mathbf{Z})}{h(t|\mathbf{Z}^*)} = \frac{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k)}{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k^*)} = \text{exp}[\sum_{k=1}^p \beta_k(Z_k-Z_k^*)] $$

-   $\mathbf{Z}$ and $\mathbf{Z}^*$ are covariate values [@klein2005]

**Kaplan-Meier**

Before generating the CPH model, it is beneficial to look at the
Kaplan-Meier curve. The Kaplan-Meier estimate ...

**CPH Model**

To generate the CPH model, a survival object must first be created.
Using the R package `survival`, the suvival formula can be generated by
running `Surv(time, event) ~ x`. That formula is then applied to the R
function `coxph(formula, data=)` which will generated the CPH model
[@R-base].

@@

For the CPH model to be valid, the following assumptions must be met. If
they are not met, the data must either be transformed.

**Proportional Hazards Assumption**

The proportional hazards assumption has been proven to be critical to
evaluate prior to using CPH when modeling survival data. The
proportional hazards assumption states that the ratio of hazard rates
for any two individuals must be constant over time regardless of the
point in time of the measurement [@Bustan2018] . This means the effect
of the covariates must be constant over time. This assumption can be
tested using the R package `survival`. The package contains a
proportional hazards test called `cox.zph()` [@R-base]. Using the
hypothesis,

@@

the `cox.zph()` calculates the scaled Schoenfeld partial residuals and
returns the $\chi^2$ value, the degrees of freedom, and p-values for
each individual covariate along with a global value. Using an
$\alpha = 0.05$ and the global p-value, the determination of whether to
reject or fail to reject $H_0$ can be made [@nahhas2025]. Individual
plots for each covariate can then be generated which provide a visual
interpretation of the p-values generated from `cox.zph()`.

**Linearity Assumption**

@@

@@

**Independent Censoring Assumption**

@@

**Independence Assumption**

@@

**Predictors**

Hypothesis test and wording for the significance of the predictors.

@@

To evaluate the model, the Cox regression concordance index will be
used. Two observed pairs are said to be concordant if an event that is
expected to occur before another does occur first. For example, if two
engines are selected from the data and the model predicts one of them to
be at a higher risk of failure than the other and the one with the
higher risk fails first, then the pairs are concordant. If the
alternative happens where the lower risk engine experiences failure
first, then the pairs are discordant.

@@

### Predictions

*Notes:*

*How coxph was created and creator what it was made for. it is semi
parametric regression*

*explain mathematically with equation*

*How data needs to be formatted - time + event and covariates*

*how it works*

*assumptions - how to test assumptions*

-   *Detail the models or algorithms used.*

-   *Justify your choices based on the problem and data.*

@@

```{r, echo = FALSE}
#packages
library(tidyverse)
library(gridExtra)
library(survival)
library(survminer)
library(ggpubr)
library(magrittr)
library(ggfortify)
library(knitr)

@@

```{r, echo=FALSE}
Histogram <- ggplot(data = distributions,
       aes(x = time)) +
  geom_histogram(binwidth = 5,
                 fill = "grey")+
  labs(x = "Survival Time",
       y = "Frequency")+
  theme_bw()

box <- ggplot(data = distributions,
                  aes(y = time))+
  geom_boxplot(fill = "gray",
               color = "black")+
  theme_bw()

@@

@@

```{r}
cox <- coxph(Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5,
             data = train)

@@

>>>>

### Data Structure

In order to use CPH to model time to event survival data, the data must contain a value for time, a binary field indicating the event being modeled, an id field, and one or multiple covariate fields. The time value is typically formatted as a time series for each individual in the dataset. Individuals will have a time value from the beginning of the observation period, to an arbitrary time that may or may not be the time the event being modeled for occurs. Data is considered censored when the individual is not given the opportunity to experience the event in the observation period and is handled differently by the model. The covariate fields in the dataset can be categorical or numerical. [@nahhas2025].

### Creating Cox Proportional Hazards Model

#### Choosing covariates

#### Generating Model
To generate the CPH model with properly formatted survival data, a survival object must first be created using the R package `survival`. The survival object is created by running `Surv(time, event)` where `time` is the time field and `event` is the event field.

To create the CPH model, the survival object is then applied to the R
function `coxph(survival object ~ x, data)` where `x` represents one or multiple covariate(s) and `data` is the survival data that is being modeled.[@R-base].

During the analysis, multiple models will be created with a different combination of covariates.

@@

When creating a CPH model, to ensure an accurate model is created the following assumptions must be met. If
they are not met, the data must either be transformed, stratified, or other covariates may need to be selected for the model. The assumptions associated with the CPH model are defined below.

#### Proportional Hazards Assumption

The proportional hazards assumption has been proven to be critical to
evaluate prior to using CPH when modeling survival data. The
proportional hazards assumption states that the ratio of hazard rates
for any two individuals must be constant over time regardless of the
point in time of the measurement. This means the effect
of the covariates must be constant over time. [@Bustan2018]

This assumption can be tested using the R package `survival`. The package contains a
proportional hazards test called `cox.zph()` [@R-base]. Using the
hypothesis,

@@

the `cox.zph()` calculates the scaled Schoenfeld partial residuals and
returns the $\chi^2$ value, the degrees of freedom, and p-values for
each individual covariate along with a global value. Using an
$\alpha = 0.05$ and the global p-value, the determination of whether to
reject or fail to reject $H_0$ can be made [@nahhas2025].

Individual plots for each covariate are generated which provide a visual
interpretation of the p-values generated from `cox.zph()`. If there is no clear slope in the residuals on the plot, it signifies further confirmation the proportional hazards assumption is met.

#### Linearity Assumption

@@

Residuals can first be found by using the `resid(type = "martingale")` function and passing martingale in the type argument creating a vector of residuals that can be plotted against the range of values for each covariate.

@@

#### Independent Censoring Assumption

@@

#### Independence Assumption

@@

@@

To evaluate the model, the Cox regression concordance index can be
used. Two observed pairs are said to be concordant if an event that is
expected to occur before another does occur first. For example, if two
engines are selected from the data and the model predicts one of them to
be at a higher risk of failure than the other and the one with the
higher risk fails first, then the pairs are concordant. If the
alternative happens where the lower risk engine experiences failure
first, then the pairs are discordant.

@@

### Model Applications and Predictions

Cox proportional hazards models can produce an assortment of useful metrics to interpret the model. The model created can be used to estimate risk of the event occurring, rate of occurrence, survival probability, and time to event.

#### Hazard Ratio

CPH models focus on providing a relative risk of the event happening based on the covariates and value for time. The hazards ratio is used to evaluate this level of risk and is calculated by finding the ratio between two individuals where one is the baseline value of a covariate at time $t$, and the other is the value of a covariate with a one unit increase. The hazards ratio is represented by the formula below.

$$ \frac{h(t|\mathbf{Z})}{h(t|\mathbf{Z}^*)} = \frac{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k)}{h_0(t)\text{exp}(\Sigma_{k=1}^{p} \beta_kZ_k^*)} = \text{exp}[\sum_{k=1}^p \beta_k(Z_k-Z_k^*)] = \text{exp}(\sum_{k=1}^{p} \beta_kZ_k) $$

-   $\mathbf{Z}$ and $\mathbf{Z}^*$ are covariate values.

A hazards ratio can be interpreted as:
$h(t|\mathbf{Z})$ < 1 indicates a decreased baseline risk as compared to the baseline.
$h(t|\mathbf{Z})$ > 1 indicates an increased baseline risk as compared to the baseline.

Hazard ratios can be interpreted to give an estimation of the event occurring. For example, a hazard ratio of 1.3 signifies that for a one unit increase in the covariate, a 30% increase in risk of the event occurring is expected. Hazard ratios can also be used to calculate hazard rates, which have theiur own applications and interpretations when modeling survival data. [@klein2005]

#### Hazard Rate

The hazard rate is defined as:

$$ h(t|\mathbf{Z}) = h_0(t)\text{exp}({\sum_{k=1}^{p} \beta_kZ_k}) $$

-   $h(t|\mathbf{Z})$ is the hazard rate at time $t$ for an individual
    with risk vector $\mathbf{Z}$

-   $\text{exp}(\sum_{k=1}^{p} \beta_kZ_k)$ is the hazard ratio as explained above.

-   $h_0(t)$ is a baseline hazard rate indicating the hazard rate when all covariates are zero.

The resulting hazard rate is a measure of the instantaneous risk of the event happening at a specific point in time. Similar to the hazard ratio, hazard rates are interpreted as:

$h(t|\mathbf{Z})$ < 1 indicates a lower risk of the event occurring at the specific point in time.
$h(t|\mathbf{Z})$ > 1 indicates a higher risk of the event occurring at the specific point in time.

Hazard rates can be used as an additional measure of risk, as well as to calculate the cumulative hazard for an individual. [@klein2005]

The hazard rate can be calculated in R by using `stats::predict()` after creating a CPH model. This function will be used to create a column showing the hazard rate for each row in the testing data.

#### Cumulative Hazard

Cumulative hazard represents the total accumulated risk of an individual for the duration of the observation period at time $t$ and is defined as:

$H(t) = \int_0^t h(t) dt$
where,
-   $H(t)$ is the cumulative hazard at time $t$

-   $h(t)$ is the hazard rate at time t

As defined by the formula, cumulative hazard is a summation of the hazard rates for an individual from time 0 to time $t$. A higher cumulative hazard indicates a higher risk of event. Cumulative hazard is used as an additional measure for risk, but it is mainly used to calculate survival probability to plot against time on survival curves.

The cumulative hazard will be calculated in R by grouping the data by the id field and finding the summation of the hazard rates for the individual through the duration of the observation period. A column will be added showing the cumulative hazard for each individual at ach point in time.

#### Calculating Survival Probability

$S(t) = exp(-H(t))$

-   $S(t)$ is the probability of survival for an individual at time $t$
-   $H(t)$ is the cumulative hazard for an individual at time $t$

The resulting survival probability will be between 0 and 1 with a value close to one indicating a high probability of the individual not experiencing the event. Conversely, a value close to 0 indicates a higher probability of the individual experiencing the event. A survival probability of .05 indicates that only 5% of individuals under the same conditions would survive. Survival probability is an important metric when modeling, visualizing, and making predictions with survival data.

Survival probabilities can be calculated in R by using `stats::predict()` after creating a CPH model. This function will be used to create a column showing the survival probability for each row in the testing data.

#### Predicting Time Until Event

### Model Visualization

There are multiple different methods to visualize data when creating a  Cox proportional hazards model. Survival curves are a popular way to show survival probabilities over time, but there are other visualizations that are helpful when interpreting the results of the model.

#### Forest Plot
Forest plots can be used to visualize the effects of each covariate in the model on the hazard ratio. The resulting plot contains each covariate on the y axis, with the estimated effect on the hazard ratio on the the x axis. The plot contains whiskers demonstrating a 95% confidence interval for the effect. A positive effect indicates a positive correlation with the hazard with a higher number signifying a more severe effect on the hazard ratio.

A forest plot will be created using the `plot_model()` function from the sjPlot package. The CPH model can be passed as the sole argument to visualize the hazard ratios.

#### Kaplan-Meier Survival Curve

After generating the CPH model, a Kaplan-Meier survival curve can be generated to view survival probabilities against time. These plots are important when modeling survival data because they clearly demonstrate the time where the event being modeled for is expected to occur.

The survival curve will be plotted by using the survival package's `survfit()` function. The object created will be plotted using `plot()`

@@

```{r, echo = FALSE}
#packages
library(tidyverse)
library(gridExtra)
library(survival)
library(survminer)
library(ggpubr)
library(magrittr)
library(ggfortify)
library(knitr)
library(sjPlot)

@@

```{r, echo=FALSE}
Histogram <- ggplot(data = distributions,
       aes(x = time)) +
  geom_histogram(binwidth = 5,
                 fill = "grey")+
  labs(x = "Survival Time (Iterations)",
       y = "Frequency")+
  theme_bw()

box <- ggplot(data = distributions,
                  aes(y = time))+
  geom_boxplot(fill = "gray",
               color = "black")+
  ylab("Time (Iterations)")+
  theme_bw()

@@

Below is a univariate test to determine which covariates to use in the model

@@

```{r}
cox <- coxph(Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5,
             data = train,
             id = id)

@@

Create df with max time for each individual to use in predictions

```{r}
predictiondata <- test %>%
  group_by(id) %>%
  slice_max(time) %>%
  select(id, time, os1, os3, sm1, sm2, sm5, status)

coef <- data.frame(Variable = c("Z1", "Z2", "Z3", "Z4", "Z5"),
           CoefficientValue = cox$coefficients)
kable(coef)
```

Hazard Rate Calculation: $h(t|Z) = h_0(t) * e^{B_1Z_1 + B_2Z_2 + B_3Z_3+B_4Z_4+B_5Z_5}$

```{r}
test$hazardrates <- predict(cox,
        newdata = test,
        type = "risk")

$H(t) = \int_0^t h(t) dt$
where,
$H(t)$ is the cumulative hazard from time 0 to time $(t)$
$h(t)$ is the hazard rate at time $t$

```{r}
test <- test %>%
  arrange(id, time) %>%
  group_by(id) %>%
  mutate(cumulative_hazard = cumsum(hazardrates))
```

Predicting survival probability

```{r}
test$survprob <- predict(cox, type = "survival", newdata = test)

Forest Plot

```{r}
plot_model(cox,dot.size = 4,line.size = 2,colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")
```
survival curve

```{r}
survfit <- survfit(cox, newdata = test)
plot(survfit, 
     col = "blue", 
     conf.int = TRUE, 
     mark.time = FALSE,
     xlab = "Time (Iterations)",
     ylab = "Survival Probability",
     main = "Survival Probability Over Time")
```

Probability Calculation: $S(t) = e^{-H(t)}$
$H(t)$ is the cumulative hazard up to time $t$
